\documentclass{article}

% if you need to pass options to natbib, use, e.g.:
%     \PassOptionsToPackage{numbers, compress}{natbib}
% before loading neurips_2026

% The authors should use one of these tracks.
% Before accepting by the NeurIPS conference, select one of the options below.
% 0. "default" for submission
% \usepackage{neurips_2026}
% the "default" option is equal to the "main" option, which is used for the Main Track with double-blind reviewing.
% 1. "main" option is used for the Main Track
%  \usepackage[main]{neurips_2026}
% 2. "position" option is used for the Position Paper Track
%  \usepackage[position]{neurips_2026}
% 3. "eandd" option is used for the Evaluations & Datasets Track
 % \usepackage[eandd]{neurips_2026}
% 4. "creativeai" option is used for the Creative AI Track
%  \usepackage[creativeai]{neurips_2026}
% 5. "sglblindworkshop" option is used for the Workshop with single-blind reviewing
 % \usepackage[sglblindworkshop]{neurips_2026}
% 6. "dblblindworkshop" option is used for the Workshop with double-blind reviewing
%  \usepackage[dblblindworkshop]{neurips_2026}

% After being accepted, the authors should add "final" behind the track to compile a camera-ready version.
% 1. Main Track
 % \usepackage[main, final]{neurips_2026}
% 2. Position Paper Track
%  \usepackage[position, final]{neurips_2026}
% 3. Evaluations & Datasets Track
 % \usepackage[eandd, final]{neurips_2026}
% 4. Creative AI Track
%  \usepackage[creativeai, final]{neurips_2026}
% 5. Workshop with single-blind reviewing
%  \usepackage[sglblindworkshop, final]{neurips_2026}
% 6. Workshop with double-blind reviewing
%  \usepackage[dblblindworkshop, final]{neurips_2026}
% Note. For the workshop paper template, both \title{} and \workshoptitle{} are required, with the former indicating the paper title shown in the title and the latter indicating the workshop title displayed in the footnote.
% For workshops (5., 6.), the authors should add the name of the workshop, "\workshoptitle" command is used to set the workshop title.
% \workshoptitle{WORKSHOP TITLE}

% "preprint" option is used for arXiv or other preprint submissions
\usepackage[preprint]{neurips_2026}
\setcitestyle{round,aysep={}}

% to avoid loading the natbib package, add option nonatbib:
%    \usepackage[nonatbib]{neurips_2026}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{minted}       % python code
\usepackage{xcolor}         % colors
\usepackage{graphicx}       % \resizebox for scaling figures to column width
\usepackage{amsmath}        % for \tfrac
\usepackage{tikz}
\usetikzlibrary{positioning, arrows.meta, shapes.geometric}
\usetikzlibrary{positioning, arrows.meta}
\usepackage{color, soul}    % highlighting during editing
% figure palette (semantic colors shared by all TikZ diagrams)
\colorlet{featuregroup}{blue!15}          % Feature Group / Concept Instance fill
\colorlet{context}{red!15}      % Context fill
\colorlet{transform}{yellow!30}           % Transform fill
\definecolor{transformedge}{HTML}{9900CC} % Transform (dashed) edges
\colorlet{concept}{green!15}              % Concept fill
\colorlet{conceptedge}{green!55!black}    % Concept link lines

% code style
\setminted[python]{
    frame=lines,
    framesep=2mm,
    baselinestretch=1.2,
    bgcolor=lightgray!20,
    fontsize=\footnotesize,
    linenos
}

% Note. For the workshop paper template, both \title{} and \workshoptitle{} are required, with the former indicating the paper title shown in the title and the latter indicating the workshop title displayed in the footnote. 
\title{Emergent Ontologies for Continuous Online Learning}


% The \author macro works with any number of authors. There are two commands
% used to separate the names and addresses of multiple authors: \And and \AND.
%
% Using \And between authors leaves it to LaTeX to determine where to break the
% lines. Using \AND forces a line break at that point. So, if LaTeX puts 3 of 4
% authors names on the first line, and the last on the second line, try using
% \AND instead of \And before the third author name.


\author{%
  Adam Powers \\
  El Dorado Hills, CA 95762 \\
  \texttt{apowers@ato.ms} \\
  % examples of more authors
  % \And
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
  % \AND
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
  % \And
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
  % \And
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
}

\begin{document}


\maketitle


\begin{abstract}
  % TODO
  TODO
\end{abstract}



\section{Introduction}
There is growing recognition that AI systems need structured world knowledge -- representations that support not just pattern matching but prediction, planning, reasoning, and transfer. Recent research in world models \hl{...}, neurosymbolics \hl{...}

Disposition of problem and existing work \hl{...}
\hl{**FEP / predictive coding:** Friston (2010; 2006); Rao \& Ballard (1999); Knill \& Pouget (2004); Friston, *Predictive coding under the free-energy principle*, 2009; FEP precision/neuromodulation literature, e.g. cholinergic-modulation PMC4235126}

In this paper we propose a cognitive architecture that is a synthesis of existing research. The architecture creates a graph-based world model that we refer to as an ontology. The ontology is created from perceptions of the environment in the same spirit as Barsalou's Perceptual Symbol Systems \citep{Barsalou_1999}. The ontology does not use first order logic for edges / relationships, but probabilistic links that approximate Friston's Free Energy Principle \citep{friston_free_2006}. The ontology is formed through one-shot acquisition, and probabilistic relationships are sculpted over time to provide higher confidence about the environment.

The foundational principles of the architecture are:
\begin{itemize}
    \item Symbols (referred to as Concepts) are required for transfactuality and one-shot learning
    \item Concepts and predictions are acquired on first encounter, confidence and precision is sculpted over time
    \item Concepts are the locus of transfactuality and prediction (as opposed to many reinforcement learning systems where states are the locus of transfactuality)
    \item Predictions must be context specific; depending on which concepts are currently present, predictions will be different
    \item Must be able to remember specific instances (where did I leave my keys) independent from generalities
    \item Concepts are assumed to be true, and truer the more they are predictively accurate, until a new experience proves otherwise. This "bold prediction" is enabled by graph structure, not probability.
    \item The structure of the ontology is emergent and stratified. There is no presupposed hierarchy, classes, or relationships. All are learned and path-specific.
    \item Prediction primacy, reward secondary (based on latent learning and sparse rewards)
    \item The goal of the system is to provide accurate predictions for homeostasis and simulation to resolve sensory valance issues that arise from time to time.
    \item Feature extractors and saliency reduce the environment to a tractable, structured space
    \item Perceptual grounding with a minimal innate endowment
    \item Must be computationally plausible on commercially available hardware in near real-time (seconds, subsecond), even if it requires high parallelism. This is to focus our work on practical solutions rather than pure theoretical approaches.
    \item The system is built out of a minimal number of mechanisms
\end{itemize}

We argue that the capacity for one continual unsupervised online learning — acquiring and immediately using new world knowledge from single experiences — does not require scale, supervision, or hand-engineered ontologies, but follows naturally when concepts are allowed to emerge bottom-up as a stratified graph structure built directly from perceptual features and sensory valence signals. We position this as a viable and underexplored alternative to the dominant approaches: large-corpus learning, world models, symbolic knowledge engineering, and prior cognitive architectures.

Not a single new mechanism but a synthesis. \hl{...}

The rest of the paper is organized as follows. \hl{...}

\section{Positioning Versus Other Approaches}
- vs. LLMs / deep learning: enables one-shot / few-shot learning from experience; does not require massive training datasets or millions of cycles
- vs. manual knowledge engineering (Cyc, OO-MDP): concepts are learned bottom-up from perception and valence signals, not hand-crafted by engineers
- vs. formal logic systems (Cyc): does not require first-order logic representations; concepts are grounded in perception
- vs. traditional RL: interposes a conceptual layer that enables transfer, abstraction, and planning beyond raw state-action mappings
- [?] Position against other cognitive architectures (ACT-R, SOAR, CLARION)?

\section{Knitting: Formation of Data Structures}
At the core is creating data structures that can be used for understanding, navigating, and manipulating the world -- especially one-shot, online learning. Environmental features and reinforcement signals are turned into concepts and temporal timelines that can be used to model the world, predict action outcomes, and simulate the environment. These concepts and temporal timelines are emergent and stratified and created from the environment itself; they are not \emph{a priori}.

One of the largest challenges with creating a world representation using data is the cold start problem -- how is the agent productive when little or no initial data is available. Therefore, the sections below are presented in the order of bootstrapping the system, putting aside the optimizations and additional benefits of having access to established concepts.

\subsection{Sensory Input and Feature Extraction}
This section lays out the basic units for forming concepts. Concepts are initially built from features, bound together at a specific space and time by their co-occurrenced. The mechanisms described in this section are well established both biologically and algorithmically and assumed to be non-controversial; however, the novel contribution is the shape of the data at the end of the feature extraction process that will be used as the basis for formulating and using concepts.

In order for the agent to reduce the data received from the environment to a manageable stream of information, multisensory input is initially broken down into a series of features using modality-specific feature extractors: edges, lines, vertices, colors, gradients, motion, in vision; frequency, harmonicity in the auditory system; edges, points, curvature, vibration in haptics; and so forth. Extracted features may be compounded to create increasingly complex features, such as creating 'T' or 'L' shape feature detectors from a combination of lines detected by the visual system. Each of these becomes a standalone feature -- an atomic node that will be used to create concepts. The modality, quantity, and types of feature extractors is not central to this argument. It is sufficient to say that there are a large number of them and the more that are available, the better the agent will be at creating differentiated concepts. \citep{pashler_attention_2016} \hl{other citations}

Each feature is mapped to a specific location in space, relative to the agent. The dimensionality is unimportant, only that features can be colocated at a specific point in time so that they can be bound together. Modality-specific saliency mechanisms \hl{Itti and Koch; Auditory saliency} (such as brightness, loudness, or motion) are used to select candidate feature groups, inhibition of return \hl{citation} is applied to ensure the agent does not become locked to a single salient point to the exclusion of all others, and cross-modal attention selects which features receive bottom-up attention using a winner-take-all mechanism. For the moment, we will put aside other context-specific mechanisms that may modify the selection of feature groups from the environment, such as agentic context that influences the selection of features. The winning features are bound together based on their spatiotemporal locus \citep{Kahneman_Treisman_Gibbs_1992, pylyshyn_role_1989, Wolfe_2021, treisman_binding_1996}  Feature binding, attention and object perception. The result is a sequential stream of feature groups at specific spatiotemporal locations. In our system, this information will be represented as a graph: a Feature Group node, with edges describing the features that create the composition of the features at that point in time.

- These "concept instances" are important because a single concept may be present multiple times at the same time. If I am holding three marbles, they all have the same concept / predictive capability, even though there are three distinct objects.

\begin{figure}[ht]
  \centering
  \begin{tikzpicture}[
    every node/.style={circle, draw, fill=featuregroup, minimum size=1.4cm, font=\small, align=center},
    edge from parent/.style={draw, {Diamond[length=6pt, width=5pt]}-, thick, color=black!70},
    level 1/.style={sibling distance=2.5cm, level distance=2cm}
  ]
    \node {Concept\\Instance}
      child {node {Feature}}
      child {node {Feature}}
      child {node {Feature}};
  \end{tikzpicture}
  \caption{A concept instance binds together the features detected at a spatiotemporal locus.}
  \label{fig:concept-instance}
\end{figure}

\subsection{Point in Time Relationships}
As this stream of feature groups is created, all pairwise relationships are defined within some window of time along multiple dimensions. For example, depending on the agent, the window of time may be seconds or milliseconds where all the feature groups are connected by relationships of distance, change in distance, orientation, and change in orientation in egocentric and allocentric frames of reference \citep{Hafri_Firestone_2021,vo_reading_2019}. These relationships are used to construct mereotopological types of concepts (e.g. part of, connected to, separate from) as well as concepts used for reasoning about relationships between disjointed objects (e.g. getting closer, touching). The relationships between these objects create Contexts, which can be used to encode specific points of time in episodic memory -- the set of features that was perceived at a specific point in time and the relation to each other, in such a way that a point in time can be reconstructed (albeit in a lossy manner). This also holds future utility in identifying and reusing instance data -- contexts can be used to identify similar situations that have occurred before.

\begin{figure}[ht]
  \centering
  \resizebox{\linewidth}{!}{%
  \begin{tikzpicture}[
    circ/.style={circle, draw, fill=featuregroup, minimum size=1.4cm, font=\small, align=center, inner sep=1pt},
    rel/.style={diamond, draw, fill=context, font=\scriptsize, align=center, inner sep=1pt},
    agg/.style={{Diamond[length=7pt, width=6pt]}-, thick, color=black!70},
    assoc/.style={-{Stealth[length=6pt]}, thick, color=black!70},
    lbl/.style={font=\scriptsize, fill=white, inner sep=1.5pt, align=center},
    x=1cm, y=1cm
  ]
    % concept instances (bottom row)
    \node[circ] (ciL) at (0,0) {Concept\\Instance};
    \node[circ] (ciR) at (9,0) {Concept\\Instance};
    % relation types
    \node[rel] (dr)  at (2.5,2.6) {Distance\\Relation\\size = 7};
    \node[rel] (dcr) at (6.5,2.6) {Distance\\Change\\Relation\\size = 2};
    % context
    \node[rel] (rg) at (4.5,5.0) {Context};

    % association: each relation type relates the two concept instances
    \draw[assoc] (dr)  -- (ciL) node[lbl, pos=0.30] {distance};
    \draw[assoc] (dr)  -- (ciR) node[lbl, pos=0.12] {distance};
    \draw[assoc] (dcr) -- (ciL) node[lbl, pos=0.12] {distance\\change};
    \draw[assoc] (dcr) -- (ciR) node[lbl, pos=0.30] {distance\\change};

    % aggregation: the context is composed of its relation types
    \draw[agg] (rg) -- (dr)  node[lbl, pos=0.6] {relation type};
    \draw[agg] (rg) -- (dcr) node[lbl, pos=0.6] {relation type};
  \end{tikzpicture}%
  }
  \caption{A Context captures the pairwise relationships -- such as distance and change in distance -- between concept instances at a single point in time. Each relation type relates the concept instances it connects. (The features associated with the concept instances have been omitted for clarity.}
  \label{fig:context}
\end{figure}

\subsection{Episodic Timeline}
Whereas the Contexts are a point-in-time snapshot of the environment, the Episodic Timeline is a series of relationships between Contexts to create a dynamic, ongoing view of the environment and its changes. The Episodic Timeline and its Transforms are the foundation for one-shot online learning. It is a recording of what has happened so that it can be used in future predictions. This timeline is not created at regular time intervals, but at the cadence that new relationships are created, which is dictated by salience; the relationships are merely before / after, not a specific amount of time before or after.

This system links Feature Groups across Contexts, identifying the same or similar feature groups in successive Contexts. This similarity is aided by feature extractors that detect motion, which reduces the dependence on similarity for tracking feature groups over time \citep{kalal_tracking-learning-detection_2012}. Each Feature group that is determined to be the same across multiple steps in the Episodic Timeline is linked back to a Concept node in the graph. The system also creates transforms between each successive Feature Group, indicating how the Feature Group changed at different points in time. These transforms and concepts are used later for prediction and simulation.

\begin{figure}[ht]
  \centering
  \resizebox{\linewidth}{!}{%
  \begin{tikzpicture}[
    circ/.style={circle, draw, fill=featuregroup, minimum size=1.4cm, font=\small, align=center, inner sep=1pt},
    rel/.style={diamond, draw, fill=context, font=\scriptsize, align=center, inner sep=1pt},
    trans/.style={rectangle, draw, fill=transform, font=\scriptsize, align=center, inner sep=3pt, minimum height=0.8cm, minimum width=1.7cm},
    assoc/.style={-{Stealth[length=6pt]}, thick, color=black!70},
    epi/.style={-{Stealth[length=9pt]}, line width=1.1pt, color=black},
    tedge/.style={-{Stealth[length=6pt]}, thick, dashed, color=transformedge},
    lbl/.style={font=\scriptsize, fill=white, inner sep=1.5pt, align=center},
    x=1cm, y=1cm
  ]
    % feature groups (middle row)
    \node[circ] (fg1)  at (0,0)    {Feature\\Group};
    \node[circ] (fg2)  at (2.2,0)  {Feature\\Group};
    \node[circ] (fg3)  at (4.4,0)  {Feature\\Group};
    \node[circ] (fg4)  at (8.2,0)  {Feature\\Group};
    \node[circ] (fg5)  at (10.4,0) {Feature\\Group};
    \node[circ] (fg6)  at (12.6,0) {Feature\\Group};
    \node[circ] (fg7)  at (14.8,0) {Feature\\Group};
    \node[circ] (fg8)  at (18.6,0) {Feature\\Group};
    \node[circ] (fg9)  at (20.8,0) {Feature\\Group};
    \node[circ] (fg10) at (23.0,0) {Feature\\Group};
    \node[circ] (fg11) at (25.2,0) {Feature\\Group};
    \node[circ] (fg12) at (27.4,0) {Feature\\Group};
    \node[circ] (fg13) at (31.2,0) {Feature\\Group};
    \node[circ] (fg14) at (33.4,0) {Feature\\Group};

    % contexts (top row) on the episodic timeline
    \node[rel] (rg1) at (2.2,3.2)  {Context};
    \node[rel] (rg2) at (11.5,3.2) {Context};
    \node[rel] (rg3) at (23.0,3.2) {Context};
    \node[rel] (rg4) at (32.3,3.2) {Context};
    \node[font=\Large] (eb) at (-2.0,3.2) {$\cdots$};
    \node[font=\Large] (ea) at (35.8,3.2) {$\cdots$};


    % transforms (bottom row)
    \node[trans] (t1) at (5.2,-3.2)  {Transform};
    \node[trans] (t2) at (9.6,-3.2)  {Transform};
    \node[trans] (t3) at (16.0,-3.2) {Transform};
    \node[trans] (t4) at (21.0,-3.2) {Transform};
    \node[trans] (t5) at (25.5,-3.2) {Transform};
    \node[trans] (t6) at (29.0,-3.2) {Transform};

    % episodic timeline (thick before/after links)
    \draw[epi] (eb)  -- (rg1);
    \draw[epi] (rg1) -- (rg2) node[lbl, pos=0.5] {episodic};
    \draw[epi] (rg2) -- (rg3) node[lbl, pos=0.5] {episodic};
    \draw[epi] (rg3) -- (rg4) node[lbl, pos=0.5] {episodic};
    \draw[epi] (rg4) -- (ea);

    % relation edges: each Feature Group relates into its Context
    \draw[assoc] (fg1)  -- (rg1) node[lbl, pos=0.28] {relation};
    \draw[assoc] (fg2)  -- (rg1) node[lbl, pos=0.28] {relation};
    \draw[assoc] (fg3)  -- (rg1) node[lbl, pos=0.28] {relation};
    \draw[assoc] (fg4)  -- (rg2) node[lbl, pos=0.28] {relation};
    \draw[assoc] (fg5)  -- (rg2) node[lbl, pos=0.28] {relation};
    \draw[assoc] (fg6)  -- (rg2) node[lbl, pos=0.28] {relation};
    \draw[assoc] (fg7)  -- (rg2) node[lbl, pos=0.28] {relation};
    \draw[assoc] (fg8)  -- (rg3) node[lbl, pos=0.28] {relation};
    \draw[assoc] (fg9)  -- (rg3) node[lbl, pos=0.28] {relation};
    \draw[assoc] (fg10) -- (rg3) node[lbl, pos=0.28] {relation};
    \draw[assoc] (fg11) -- (rg3) node[lbl, pos=0.28] {relation};
    \draw[assoc] (fg12) -- (rg3) node[lbl, pos=0.28] {relation};
    \draw[assoc] (fg13) -- (rg4) node[lbl, pos=0.28] {relation};
    \draw[assoc] (fg14) -- (rg4) node[lbl, pos=0.28] {relation};

    % transform edges: earlier Feature Group -> Transform -> later Feature Group
    \draw[tedge] (fg1)  -- (t1);  \draw[tedge] (t1) -- (fg5);
    \draw[tedge] (fg3)  -- (t2);  \draw[tedge] (t2) -- (fg7);
    \draw[tedge] (fg6)  -- (t3);  \draw[tedge] (t3) -- (fg10);
    \draw[tedge] (fg4)  -- (t4);  \draw[tedge] (t4) -- (fg13);
    \draw[tedge] (fg8)  -- (t5);  \draw[tedge] (t5) -- (fg14);
    \draw[tedge] (fg11) -- (t6);  \draw[tedge] (t6) -- (fg13);
  \end{tikzpicture}%
  }
  \caption{The Episodic Timeline links successive Contexts -- the point-in-time snapshots of Figure~\ref{fig:context} -- with before/after \emph{episodic} edges, forming an ongoing record of the environment. Each Context relates the Feature Groups present at that point in time, and each \emph{Transform} captures how a Feature Group changes between one point in time and a later one. (The ``before'' and ``after'' ellipses indicate that the timeline extends indefinitely in both directions.)}
  \label{fig:episodic-timeline}
\end{figure}

\subsection{Prediction}
Prediction starts with identifying similar Concepts and Contexts. The worst case scenario is the agent being presented an environment with no prior context. In this scenario Concepts are identified starting with saliency mechanisms selecting a Feature Group and then matching the Feature Group to a Concept based on spreading activation of features in Feature Groups -- whichever Concept has the most features in common with the saliency selected feature group is the most likely matching concept. This happens iteratively as saliency shifts attention around the environment. As Concepts are identified in the current environment, their relationships start to come into play, using Contexts associated with the Feature Groups of the concept to identify Feature Groups that have co-occurred in previous experience. Finally, as time moves forward and motion and change comes into play, the Transformations between Feature Groups associated with a concept can remove any final ambiguity if there are multiple potentially matching concepts or if the overall similarity based on Features and Contexts alone was low.

Note that the Episodic Timeline graph of the previous section can be refactored to be Concept-centric. Each concept has its own Feature Groups, Transforms from one set of features to another, and the Contexts under which those transforms occurred. Having recognized Concepts of the scene, their current relationships (if any) can be used to predict the next Transform that will occur, which can be applied to the current world model of the agent to predict the next state. This occurs across all identified Concepts, predicting the next state of the environment.

After matching the current context, or the concepts of the context if no context exists, prediction uses these edges to determine the most likely transforms that will apply to each concept and predict the next state of each concept in terms of its features and relationships to other concepts. The Normal-Gamma conjugate prior has its typical form:

\begin{figure}[ht]
  \centering
  \begin{tikzpicture}[
    conc/.style={circle, draw, fill=concept, minimum size=1.4cm, font=\small, align=center, inner sep=1pt},
    circ/.style={circle, draw, fill=featuregroup, minimum size=1.4cm, font=\small, align=center, inner sep=1pt},
    rel/.style={diamond, draw, fill=context, font=\scriptsize, align=center, inner sep=1pt},
    trans/.style={rectangle, draw, fill=transform, font=\scriptsize, align=center, inner sep=3pt, minimum height=0.8cm, minimum width=1.7cm},
    cedge/.style={thick, color=conceptedge},
    tedge/.style={-{Stealth[length=6pt]}, thick, dashed, color=transformedge},
    assoc/.style={-{Stealth[length=6pt]}, thick, color=black!70},
    x=1cm, y=1cm
  ]
    \node[conc]  (cpt) at (0,3.2)   {Concept};
    \node[circ]  (fg1) at (-3,0)    {Feature\\Group};
    \node[circ]  (fg2) at (3,0)     {Feature\\Group};
    \node[trans] (t)   at (0,0)     {Transform};
    \node[rel]   (rg1) at (-3,-3.2) {Context};
    \node[rel]   (rg2) at (3,-3.2)  {Context};

    % concept linked to its two feature groups (green lines)
    \draw[cedge] (cpt) -- (fg1);
    \draw[cedge] (cpt) -- (fg2);
    % transform between the two feature groups
    \draw[tedge] (fg1) -- (t);
    \draw[tedge] (t) -- (fg2);
    % each feature group relates into its context
    \draw[assoc] (fg1) -- (rg1);
    \draw[assoc] (fg2) -- (rg2);
  \end{tikzpicture}
  \caption{A Concept (green) retains links to the Feature Groups that instantiate it. Two such Feature Groups are shown, together with the \emph{Transform} that captures the change between them and the \emph{Context} that each one participated in.}
  \label{fig:concept-structure}
\end{figure}

\subsection{Acquisition}
- At this point we have a stream of information, however it is too much to retain and not all of it is necessary to retain for future purposes
- Novelty creates new concepts: features, relationships, etc that are not similar to others
- Concepts are the locus of predictability and transferability. They have relationships to all their (retained) concept instances, the relationships that were present in those instances, and the transforms that occurred in those instances. This enables them to identify similar circumstances based on the other concepts present and apply the same transforms under those conditions to predict or simulate what will happen in future situations.

\subsection{Sculpting Edges}
When a concept is successfully matched, the edges that that are used for the match have their probabilities increased information and the edges that were expected but absenthave their probabilities decreased (i.e. the edges are "sculpted"). Note that inheritance relationships are not sculpted and thus aren't described in this section. See Section \ref{sec:abstraction} for more about inheritance.

There are two types of sculpted edges: a PresenceEdge that is used to indicate that the relationship is or isn't present and a ValueEdge that is used to represent both presence and the the probability distribution of a perceptual feature (e.g. the probability distribution of a certain hue or brightness for a color).

The types of relationships that are sculpted are:
\begin{itemize}
    \item Concept $\dashrightarrow$ FeatureGroup / ConceptInstance
    \item FeatureGroup $\rightsquigarrow$ Feature
    \item (Concept, Context) $\dashrightarrow$ Transform
    \item Concept $\dashrightarrow$ Context
    \item Concept $\rightsquigarrow$ ContextRelationship
    \item source FeatureGroup $\rightsquigarrow$ Transform $\rightsquigarrow$ destination FeatureGroup
    \item Context $\rightsquigarrow$ Significance
    \item source Context $\rightsquigarrow$ Action $\rightsquigarrow$ destination Context
\end{itemize}

Where $\dashrightarrow$ is a PresenceEdge and $\rightsquigarrow$ is a ValueEdge. 

PresenceEdges carry a Beta count of how often the target of the edge was present versus absent when the structure was matched. When an expected edge is present $a$ updates $a' = a+1$, and when an expected edge is absent $b$ is updated $b' = b+1$. The presence probability $a/(a+b)$ is the edge's \emph{criteriality}: how necessary the target is during the matching process. It also provides occlusion and variation robustness: an expected-but-absent relationship is not disqualifying, it contributes $1 - a/(a+b)$ to the match. This aids in feature recognition, where a feature may be occluded; and it aids in concept-context relationships where a missing concept will still yield a context that may still have relevant predictions.

ValueEdges extend the functionality of PresenceEdges. In addition to the Beta values on a PresenceEdge, ValueEdges carry a Normal-Gamma conjugate prior that is "sculpted" during the use of the edge to determine the probability of the nodes along the edge being a match. The Normal-Gamma summary takes the form $(\mu, \kappa, \alpha, \beta)$.

\begin{itemize}
    \item $\mathbf{\mu}$, the current best estimate of the feature's value for this concept (the Normal-Gamma mean);
    \item $\mathbf{\kappa}$, the *effective evidence* behind $\mu$ — its "virtual sample size" — which also sets how much a new observation moves the estimate and how strongly the edge resists being overturned;
    \item $\mathbf{\alpha}$, the Gamma shape, accumulating evidence about the feature's precision;
    \item $\mathbf{\beta}$, the Gamma rate, accumulating observed squared deviation, so that $\alpha_{NG}/\beta$ estimates the precision (tightness) of the feature for this concept.
\end{itemize}

On the n-th consistent observation x, the standard Normal-Gamma updates are

$$\mu' = \frac{\kappa\mu + x}{\kappa + 1}, \qquad \kappa' = \kappa + 1, \qquad \alpha' = \alpha + \tfrac{1}{2}, \qquad \beta' = \beta + \frac{\kappa}{\kappa+1}\cdot\frac{(x - \mu)^2}{2}.$$

See Section \ref{pseudocode:edges} for pseudocode detailing the implementation of Edges.

\begin{figure}[ht]
  \centering
  \begin{tikzpicture}[
    conc/.style={circle, draw, fill=concept, minimum size=1.5cm, font=\small, align=center, inner sep=1pt},
    feat/.style={circle, draw, fill=featuregroup, minimum size=1.5cm, font=\small, align=center, inner sep=1pt},
    chip/.style={rectangle, draw, rounded corners, fill=black!3, align=center, inner sep=5pt, minimum width=3.7cm, font=\footnotesize},
    icon/.style={rectangle, draw=black!25, rounded corners, fill=white, minimum width=3.9cm, minimum height=1.7cm},
    cedge/.style={line width=1.4pt, color=conceptedge},
    pin/.style={-{Stealth[length=5pt]}, thick, color=black!55},
    edgelbl/.style={font=\scriptsize, fill=white, inner sep=1.5pt},
    x=1cm, y=1cm
  ]
    % --- the two real nodes and the single edge between them ---
    \node[conc] (c) at (0,0)  {Concept};
    \node[feat] (f) at (11,0) {Feature};
    \draw[cedge] (c) -- (f) node[edgelbl, midway, above=1pt] {Concept $\to$ Feature edge};
    \coordinate (mid) at (5.5,0);

    % --- distribution icons (top of each column) ---
    % presence icon: skewed Beta, mass near 1 (a criterial, usually-present feature)
    \node[icon] (pi) at (3.0,-1.55) {};
    \begin{scope}[shift={(3.0,-1.55)}]
      \draw[->, black!35] (-1.55,-0.55) -- (1.7,-0.55);
      \draw[line width=1pt, conceptedge] plot[smooth] coordinates
        {(-1.5,-0.53) (-0.7,-0.4) (0,-0.12) (0.6,0.34) (1.05,0.58) (1.4,0.05)};
      \node[font=\tiny, black!55] at (-1.5,-0.78) {0};
      \node[font=\tiny, black!55] at (1.62,-0.78) {1};
    \end{scope}
    % value icon: Gaussian bell centered on mu
    \node[icon] (vi) at (8.0,-1.55) {};
    \begin{scope}[shift={(8.0,-1.55)}]
      \draw[->, black!35] (-1.6,-0.55) -- (1.7,-0.55);
      \draw[line width=1pt, conceptedge] plot[smooth] coordinates
        {(-1.4,-0.54) (-0.8,-0.34) (-0.35,0.12) (0,0.56) (0.35,0.12) (0.8,-0.34) (1.4,-0.54)};
      \draw[black!35] (0,-0.5) -- (0,-0.64) node[font=\tiny, black!55, below, inner sep=1pt] {$\mu$};
    \end{scope}

    % --- pins from the edge down to each icon ---
    \draw[pin] (mid) -- (pi.north);
    \draw[pin] (mid) -- (vi.north);

    % --- text chips (below each icon) ---
    \node[chip] (pc) at (3.0,-3.6) {\textbf{Presence}\\[1pt] $\mathrm{Beta}(a,b)$\\[2pt] $P(\mathrm{present})=\frac{a}{a+b}$\\[1pt] {\scriptsize $=$ criteriality}};
    \node[chip] (vc) at (8.0,-3.6) {\textbf{Value}\\[1pt] Normal-Gamma\\[2pt] $(\mu,\kappa,\alpha,\beta)$\\[1pt] {\scriptsize predictive: Student-$t$}};

    % --- multiply symbol between the two chips ---
    \node[font=\LARGE] at (5.5,-3.6) {$\times$};

    % --- combined likelihood (bottom) ---
    \node[rectangle, draw, rounded corners, fill=black!3, align=center, inner sep=7pt, minimum width=10.4cm, font=\small]
      (comb) at (5.5,-5.75)
      {match likelihood:\; $L \;=\; \dfrac{a}{a+b}\;\times\;\text{Student-}t\!\left(x \mid \mu,\kappa,\alpha,\beta\right)$\\[4pt]
       {\footnotesize expected but absent (occluded):\; $L = 1 - \dfrac{a}{a+b}$}};
    \draw[pin] (pc.south) -- (comb.north -| pc.south);
    \draw[pin] (vc.south) -- (comb.north -| vc.south);
  \end{tikzpicture}
  \caption{Anatomy of a ValueEdge. Every ValueEdge stores two conjugate summaries, sculpted independently: a \emph{presence} distribution (Beta counts, whose mean $a/(a+b)$ is the feature's \emph{criteriality}) and a \emph{value} distribution (Normal-Gamma). Matching multiplies the two into a single match likelihood; an expected-but-absent feature contributes $1-a/(a+b)$ rather than disqualifying the match.}
  \label{fig:edge-anatomy}
\end{figure}

\subsection{Reification}
- Uses strength of PresenceEdges (i.e. probability ~1) to create new parent Concepts
- additive reification leads to mereology relationships: features that regularly occur on the same concept, in the same relationship, and have the same transforms are part of the same Concept (e.g. fingers are part of hand because they move with the hand, have consistent relationships to the hand, and move when the hand moves so they have the same transforms).
- subtractive reification \hl{TODO}

\subsection{Sensory Valence}
- Sensory valence (reward and punishment) signals (hunger, pain, pleasure, etc.) are not binary, but are continuous (degree of hunger, degree of pain, degree of pleasure, etc.). The signal is negative for situations to be avoided.
- Sensory valence signals are nodes that are linked to contexts based on the their continuous value as a weight, enabling them to be retrieved when similar situations are encountered: contexts that were tagged with hunger can be recalled when the agent is facing the same signal, enabling it to identify contexts that may aid in the current situation.
- They may also be immediately available during recall to identify near-term objects or actions to steer towards or away from.
- Sensory valence acts as an amplification signal for concept and context acquisition, ensuring that high valence scenarios are available for recall to guide attraction or avoidance in action selection.
- Valence also acts as a multiplier on the novelty threshold: high-stakes prediction failures spawn new structure sooner than low-stakes ones.


\begin{figure}[ht]
  \centering
  \resizebox{\linewidth}{!}{%
  \begin{tikzpicture}[
    circ/.style={circle, draw, fill=featuregroup, minimum size=1.4cm, font=\small, align=center, inner sep=1pt},
    rel/.style={diamond, draw, fill=context, font=\scriptsize, align=center, inner sep=1pt},
    trans/.style={rectangle, draw, fill=transform, font=\scriptsize, align=center, inner sep=3pt, minimum height=0.8cm, minimum width=1.7cm},
    val/.style={circle, draw, fill=green!15, minimum size=1.4cm, font=\small, align=center, inner sep=1pt},
    assoc/.style={-{Stealth[length=6pt]}, thick, color=black!70},
    epi/.style={-{Stealth[length=9pt]}, line width=1.1pt, color=black},
    tedge/.style={-{Stealth[length=6pt]}, thick, dashed, color=transformedge},
    lbl/.style={font=\scriptsize, fill=white, inner sep=1.5pt, align=center},
    x=1cm, y=1cm
  ]
    % feature groups (middle row)
    \node[circ] (fg1)  at (0,0)    {Feature\\Group};
    \node[circ] (fg2)  at (2.2,0)  {Feature\\Group};
    \node[circ] (fg3)  at (4.4,0)  {Feature\\Group};
    \node[circ] (fg4)  at (8.2,0)  {Feature\\Group};
    \node[circ] (fg5)  at (10.4,0) {Feature\\Group};
    \node[circ] (fg6)  at (12.6,0) {Feature\\Group};
    \node[circ] (fg7)  at (14.8,0) {Feature\\Group};
    \node[circ] (fg8)  at (18.6,0) {Feature\\Group};
    \node[circ] (fg9)  at (20.8,0) {Feature\\Group};
    \node[circ] (fg10) at (23.0,0) {Feature\\Group};
    \node[circ] (fg11) at (25.2,0) {Feature\\Group};
    \node[circ] (fg12) at (27.4,0) {Feature\\Group};
    \node[circ] (fg13) at (31.2,0) {Feature\\Group};
    \node[circ] (fg14) at (33.4,0) {Feature\\Group};

    % contexts (top row) on the episodic timeline
    \node[rel] (rg1) at (2.2,3.2)  {Context};
    \node[rel] (rg2) at (11.5,3.2) {Context};
    \node[rel] (rg3) at (23.0,3.2) {Context};
    \node[rel] (rg4) at (32.3,3.2) {Context};
    \node[font=\Large] (eb) at (-2.0,3.2) {$\cdots$};
    \node[font=\Large] (ea) at (35.8,3.2) {$\cdots$};

    % valence tags (above the contexts)
    \node[val] (v1) at (2.2,6.2)  {Valence};
    \node[val] (v2) at (23.0,6.2) {Valence};
    \node[val] (v3) at (32.3,6.2) {Valence};

    % transforms (bottom row)
    \node[trans] (t1) at (5.2,-3.2)  {Transform};
    \node[trans] (t2) at (9.6,-3.2)  {Transform};
    \node[trans] (t3) at (16.0,-3.2) {Transform};
    \node[trans] (t4) at (21.0,-3.2) {Transform};
    \node[trans] (t5) at (25.5,-3.2) {Transform};
    \node[trans] (t6) at (29.0,-3.2) {Transform};

    % episodic timeline (thick before/after links)
    \draw[epi] (eb)  -- (rg1);
    \draw[epi] (rg1) -- (rg2) node[lbl, pos=0.5] {episodic};
    \draw[epi] (rg2) -- (rg3) node[lbl, pos=0.5] {episodic};
    \draw[epi] (rg3) -- (rg4) node[lbl, pos=0.5] {episodic};
    \draw[epi] (rg4) -- (ea);

    % relation edges: each Feature Group relates into its Context
    \draw[assoc] (fg1)  -- (rg1) node[lbl, pos=0.28] {relation};
    \draw[assoc] (fg2)  -- (rg1) node[lbl, pos=0.28] {relation};
    \draw[assoc] (fg3)  -- (rg1) node[lbl, pos=0.28] {relation};
    \draw[assoc] (fg4)  -- (rg2) node[lbl, pos=0.28] {relation};
    \draw[assoc] (fg5)  -- (rg2) node[lbl, pos=0.28] {relation};
    \draw[assoc] (fg6)  -- (rg2) node[lbl, pos=0.28] {relation};
    \draw[assoc] (fg7)  -- (rg2) node[lbl, pos=0.28] {relation};
    \draw[assoc] (fg8)  -- (rg3) node[lbl, pos=0.28] {relation};
    \draw[assoc] (fg9)  -- (rg3) node[lbl, pos=0.28] {relation};
    \draw[assoc] (fg10) -- (rg3) node[lbl, pos=0.28] {relation};
    \draw[assoc] (fg11) -- (rg3) node[lbl, pos=0.28] {relation};
    \draw[assoc] (fg12) -- (rg3) node[lbl, pos=0.28] {relation};
    \draw[assoc] (fg13) -- (rg4) node[lbl, pos=0.28] {relation};
    \draw[assoc] (fg14) -- (rg4) node[lbl, pos=0.28] {relation};

    % valence edges: each Valence node tags one Context
    \draw[assoc] (v1) -- (rg1) node[lbl, pos=0.5] {valence};
    \draw[assoc] (v2) -- (rg3) node[lbl, pos=0.5] {valence};
    \draw[assoc] (v3) -- (rg4) node[lbl, pos=0.5] {valence};

    % transform edges: earlier Feature Group -> Transform -> later Feature Group
    \draw[tedge] (fg1)  -- (t1);  \draw[tedge] (t1) -- (fg5);
    \draw[tedge] (fg3)  -- (t2);  \draw[tedge] (t2) -- (fg7);
    \draw[tedge] (fg6)  -- (t3);  \draw[tedge] (t3) -- (fg10);
    \draw[tedge] (fg4)  -- (t4);  \draw[tedge] (t4) -- (fg13);
    \draw[tedge] (fg8)  -- (t5);  \draw[tedge] (t5) -- (fg14);
    \draw[tedge] (fg11) -- (t6);  \draw[tedge] (t6) -- (fg13);
  \end{tikzpicture}%
  }
  \caption{Sensory valence signals attach to the episodic timeline of Figure~\ref{fig:episodic-timeline}: each \emph{Valence} node tags a Context with the continuous reward or punishment signal (e.g. hunger, pain, pleasure) present at that point in time. Not every Context carries a valence tag. The tags index the timeline for retrieval: when a similar signal recurs, the tagged Contexts and their concepts can be recalled.}
  \label{fig:sensory-valence}
\end{figure}

\subsection{Conceptual Recall}
- Concepts are recalled through three mechanisms: 1) temporal relationships from the current context, which is extremely fast 2) bottom-up feature recognition, which is slow but broad; this uses the features (sound frequency, lines, intersections) with spreading activation to identify a concept; 3) contexts -- based on other recognized concepts in the current context, using contexts to identify candidate concepts.

\subsection{Retention / Forgetting}
- Repetition through recall leads to retention (i.e. reinforcement). Every successful recall sculpts edges to increase their probability of their associated concepts being referenced in the future.
- Forgetting is the process of reducing the weight of edges through decay (i.e. the effective evidence $\kappa$ and the presence counts decay towards a plasticity floor). Edges are not removed, but lose their priority in recall and / or reduce their confidence. "Forgotten" concepts are hard to reach but not gone and on re-encounter they can be recalled and the edges will quickly increase in confidence to restore its utility.
- Forgetting is essential to the plasticity of the agent -- de-prioritizing old information allows new information to be utilized in the correct context. For example, if an object is moved the agent must be able to reduce the probability of the old location and increase the probability of the new location so that it can be recalled in the correct context.
- Explains both aspects of forgetting due to interference: retroactive interference occurs because the graph develops new, stronger relationships that make old relationships difficult to reach; proactive interference because the weights are already established heavily in favor of historical concepts.

\citep{ryan_frankland_2022, underwood_1957, oleary_natural_2024, bjork_bjork_1992}

\subsection{Concept Abstraction, Hoisting, Inheritance}\label{sec:abstraction}
- Frequently co-occurring and predictive features, relationships, and transforms are abstracted out to become new concepts. This is an economic savings for three reasons: 1) rather than holding on to a large amount of detail, only the necessary detail for the high-confidence predictions needs to be retained; 2) relationships can be formed and predictions can happen at higher levels, enabling faster predictions; 3) the newly extracted / abstracted concepts can be applied to more scenarios and newly encountered scenarios, enabling transference of knowledge / prediction.
- Co-activation during matching leads to abstraction and is a different mechanism than reification. Reification uses strength of sculpted edges as a grouping mechanism, not co-activation.
- When a new level of concept is created, it hoists up the common contexts, feature groups, and transforms -- adding them to the new parent concept and removing them from the common concepts.
- Child concepts inherit the contexts, feature groups, and transforms of their parents -- the common "dog" class of features and behaviors applies to all its children. Individual children can override the inheritance by retaining or adding specific features, contexts, or transforms to their own concepts enabling child-concept specific predictions. For example, "all dogs are generally nice but Fido tries to bite".
- Transforms can be abstracted to concepts. This enables reasoning and transference around concepts such as "fast" and "to the left"
- The concept instances that contribute to the concept also contribute their known relationships and previous behaviors: the concept is expected to behave the same when the concepts surrounding it are the same.
- This process is iterative: concepts can be iteratively abstracted, not requiring grounded concepts to form higher-level concepts. (See iterative / recursive abstraction below)
- \hl{TODO: how do "red" (a pure feature) and "above" (a pure relationship) and "fast" (a pure transform) become concepts?}

\citep{mandler_perceptual_2000}

Concept First Layer Abstraction Diagram: \hl{TODO}
Concept Second Layer Abstraction Diagram: \hl{TODO}

\subsection{Iterative / Recursive Abstraction}
- Can abstract concepts, relationships, transforms across different types into new concepts.
- Iterative abstraction need not be across common types (concept + concept abstraction; etc.); New concepts may be formed on mixed types (concept + relationship abstraction; etc.)
- Example: \hl{TODO}

Iterative Abstraction Diagram(s): \hl{TODO}

\subsection{Prediction}
- The agent is able to predict what will happen next based on the concept instances that are present. Each concept instance, with enough previous experience, has its own predicted behavior and predicted behavior in the presence of other known concepts. 
- Since the agent can predict the currently present objects, it can predict how its own actions will influence them.
- Successful prediction leads to reinforcement / additional weighting of the concepts used in the prediction.
- Failure in prediction leads to a new temporal association with the concepts.

\subsection{Action}
\begin{itemize}
    \item Scripts of actions (a/k/a "habits") based on strongly associated conceptual timelines
    \item Deviations based on prediction error are quickly recovered
    \item Actions are selected based on prediction: each individual object is predicted giving the agent a prediction of the upcoming state. The agent selects actions by searching through the immediate predictions and potential sensory valence.
    \item Actions are recorded with relation to the current context, enabling transference of actions to future scenarios with the same or similar context.
    \item Actions are not selected based on long-term sensory valence, they are selected for general homeostasis. Only if there is a problem (predicted threat, negative sensory valence due to hunger, etc.) will they seek out solutions using Option Generation and Simulation.
\end{itemize}

\subsection{Simulation}
- Concepts and temporal traces enable simulation of known and novel situations
- Play through temporal links
- Modified simulation: play through temporal links with modifications

\subsection{Option Generation}
- Indexes: concepts, timelines, valence
- "I'm hungry" -- find solutions to the problem through association; uses valence to find previous solutions
- Seeking out solutions to current problems is based on using Sensory Valence associated with the specific anticipated problem to identify options for solutions, then simulation to select options and potential paths to the options.

\subsection{Introspection}
- Internal states must be available for observation to enable introspection
- Internal states that are available for introspection likely include confidence levels of prediction and simulation, recent memory, and recent perception -- basically the ability to inspect the recent episodic timeline.


\section{References}
\bibliographystyle{plainnat}
\bibliography{references}

\appendix
\section{Pseudocode}\label{sec:appx:a}

\subsection{Concepts}\label{pseudocode:concepts}

\subsection{Contexts}\label{pseudocode:contexts}

\subsection{Transforms}\label{pseudocode:transforms}

\subsection{Edges}\label{pseudocode:edges}
\begin{minted}{python}
class Edge:
    # the Feature that the edge connects to
    # TODO: edges aren't just for features
    feature: Feature

    # Beta counts (present, absent)
    # presence_prob = a/(a+b) = criteriality
    a: int
    b: int

    # Normal-Gamma edge probability for continuous feature values (e.g. color)
    mu: float
    kappa: float
    alpha: float
    beta: float

    # Dirichlet edge probability for categorical feature values (e.g. up / down / left / right)
    # TODO: currently not used
    dir_counts: dict[any, int]  # count of features, any is the range of the feature the edge points to
    n: int

    def get_feature(self, features: FeatureGroup) -> Feature | None:
        # looks up the Feature that corresponds to the Feature expected by this Edge

    def probability_present(self) -> float:
        self.a / (self.a + self.b)

    # is this Edge present and of the right value in a set of Features
    # TODO: should be more than features
    def likelihood(self, features: FeatureGroup) -> float:
        present: bool = self.feature in features
    
        if not present:
            # expected but absent isn't a deal-breaker, maybe an occluded feature
            return(1 - self.probability_present())

        # calculate Student-t PDF for value
        val: float = self.get_feature(features).value
        value_term: float = scipy.stats.t(val, mu=self.mu, kappa=self.kappa, alpha=self.alpha, beta=self.beta)
    
        return value_term * self.probability_present()

    def update(self, features: FeatureGroup) -> None:
        present: bool = self.feature in features

        if not present:
            self.b += 1
            return

        # update presence counts
        self.a += 1
        self.n_raw += 1

        # sculpt edge value
        val: float = self.get_feature(features).value
        k = self.kappa
        mu_old = self.mu
        self.mu = (k * mu_old + x) / (k + 1)
        self.beta = self.beta + (k / (k + 1)) * (val - mu_old)^2 / 2
        self.kappa = k + 1
        edge.alpha = self.alpha + 0.5
        
\end{minted}

\subsection{Saliency}\label{pseudocode:saliency}
\begin{minted}{python}
def get_saliency_points():
    get_vision_saliency_points()
    get_aural_saliency_points()
    get_tactile_saliency_points()
    # etc.

    crossmodal_saliency_points()
\end{minted}

\begin{minted}{python}
def get_vision_saliency_points():
    feature_map = vision_feature_extractors()
\end{minted}

\subsection{Main Loop}\label{pseudocode:main-loop}
\begin{minted}{python}
def main():
    obs = get_observation()

    # --- 1. do feature extraction
    features = feature_extraction()

    # --- 2. get the salient features, bound together by location
    feature_groups = get_saliency_points(features)

    # --- 3. resolve previous prediction

    # --- 5. identify concepts
    concepts = identify_concepts(prev_context, feature_groups)

    # --- 6. identify context

    # --- 7. predict

    # --- 8. take action
\end{minted}

\section{Sufficiency Tests}
\subsection{A cat is not a dog}
- The agent has seen multiple dogs before of different varieties: roughly large, furry, four legged creatures, that pant, and are friendly. The agent encounters a new dog that generally meets the criteria of previous dogs, but has a few subtle differences such as pointier ears and whiskers. When the agent approaches the dog, it runs away which is counter to prediction. When the agent continues to pursue the dog it hisses and swipes its claws at the agent. This is not the behavior of previous dogs, it must be something new. A cat. 
- This is a demonstration of bifurcation of concepts based on different predictions of the objects.

\subsection{There's a bear in my kitchen}
- The agent enters the kitchen in a house -- a familiar scenario. There is a bear in the kitchen. The agent is able to recognize the bear without having encountered a bear in the kitchen before. 
- This is a demonstration of bottom-up recognition with incongruent context.
\subsection{Social Transfer}
- \hl{TODO}
\subsection{Emotion}
- \hl{TODO}

\subsection{"Democracy"}
- A frequent challenge to grounded concepts is that they can't create the concept of "democracy" (as an example of a sufficiently abstract concept). Through iterative abstraction, grounded perceptual symbols would get further away from their grounding to create abstract symbols with no grounding. Specifically a high-level sketch of obtaining the concept of "democracy" would look something like:
	1. Learn the instance of "mother" as the one that is feeding the agent
	2. Learn the concept of "mothers" as other agents having a similar situation
	3. Experience "mothers" as setting and enforcing rules to create a concept of "authority figure"
	4. Learn through social transfer that there are different "authority figures", such as kings and presidents
	5. Learn that "kings" can be cruel and arbitrary and that fairness / equity / predictability is a desirable outcome
	6. Learn that people can vote to create equitable rules for the masses and that this is called "democracy"
\subsection{Language}
- Language is sounds and symbols tied to concepts, it has no special mechanisms.
- Sentence structure and grammar is learned by patterns.
- Symbol-to-concept binding is arbitrary – unlike feature-grounded concepts, a word shares no features with its referent.
- Language is also an input channel: create a token based on a sound or a set of symbols that needs to be filled in. It further gets associated with the relationships of words or concepts around it, based on explication, context clues, or assumption.
\subsection{Dog and Frisbee}
- A person throws a frisbee towards a dog. The agent is able to predict that the dog will chase towards the frisbee and whether or not it will be able to intercept it. 
- The prediction is not based on any reward signal and the agent has no reason to maintain a reward around dogs and frisbees.
\subsection{Stop signs don't matter when you are walking}
- \hl{TODO}

\subsection{The drapes are different in his memory}
- Legion season 1, they go back into the memory of Legion to his therapists office, then they visit the office in real life. The drapes have a different pattern in his memory than in real life.
- This is an example of episodic memory encoding rough concepts rather than precise feature group details.

\subsection{A handful of marbles}
- Agent has a memory of holding three marbles. In recollection it may recall the different colors and differentiation of size, but not the specific details of the patterns on the marbles. It will also recall the size of the marbles relative to each other and relative to the hand; the rough positioning in the hand; etc.
- This is an example of retaining some features in a feature group but eliminating others.

\subsection{Driving to work by mistake}
- You are running an errand on the weekend. The route you take starts by following the same route you always take to work, and then diverges part way through to the route you need to continue to run your errand. You accidentally miss your turn for your errand and continue driving to work.
- This is a demonstration of habits (strong association and strength of recall) dictating actions.

\subsection{Daydreaming}
- \hl{TODO}

\end{document}
